You can install the latest released version using the normal `R` function, though you need to point to `r-universe` instead of CRAN:

```{r, eval = FALSE}
install.packages(
  "primarycensored",
  repos = "https://epinowcast.r-universe.dev"
)
```

Alternatively, you can use the [`remotes` package](https://remotes.r-lib.org/) to install the development version from Github (warning! this version may contain breaking changes and/or bugs):

```{r, eval = FALSE}
remotes::install_github(
  "epinowcast/primarycensored",
  dependencies = TRUE
)
```

Similarly, you can install historical versions by specifying the release tag (e.g. this installs [`0.2.0`](https://github.com/epinowcast/primarycensored/releases/tag/v0.2.0)):

```{r, eval = FALSE}
remotes::install_github(
  "epinowcast/primarycensored",
  dependencies = TRUE, ref = "v0.2.0"
)
```

*Note: You can also use that last approach to install a specific commit if needed, e.g. if you want to try out a specific unreleased feature, but not the absolute latest developmental version.*
