You can install the latest released version from CRAN using the standard `install.packages` function:

```{r, eval = FALSE}
install.packages("primarycensored")
```

Alternatively, you can install the latest release from our r-universe repository:

```{r, eval = FALSE}
install.packages("primarycensored", repos = "https://epinowcast.r-universe.dev")
```

To install the development version from GitHub (warning! this version may contain breaking changes and/or bugs), use the [`pak` package](https://pak.r-lib.org/):

```{r, eval = FALSE}
pak::pak("epinowcast/primarycensored")
```

Similarly, you can install historical versions by specifying the release tag (e.g., [`v0.2.0`](https://github.com/epinowcast/primarycensored/releases/tag/v0.2.0)):

```{r, eval = FALSE}
pak::pak("epinowcast/primarycensored@v0.2.0")
```

*Note: You can also use the above approach to install a specific commit if needed, for example, if you want to try out a specific unreleased feature, but not the absolute latest developmental version.*
