---
title: "Fitting distributions using primarycensored and cmdstan"
description: "A guide on how to use primarycensored with Stan for Bayesian inference of epidemiological delay distributions."
author: Sam Abbott
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting distributions using primarycensored and cmdstan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What are we going to do in this Vignette

In this vignette, we'll demonstrate how to use `primarycensored` in conjunction with Stan for Bayesian inference of epidemiological delay distributions. We'll cover the following key points:

1. Simulating censored delay distribution data
2. Fitting a naive model using cmdstan
3. Evaluating the naive model's performance
4. Fitting an improved model using `primarycensored` functionality
5. Comparing the `primarycensored` model's performance to the naive model

## What might I need to know before starting

This vignette builds on the concepts introduced in the [Getting Started with primarycensored](primarycensored.html) vignette and assumes familiarity with using Stan tools as covered in the [How to use primarycensored with Stan](using-stan-tools.html) vignette.

## Packages used in this vignette

Alongside the `primarycensored` package we will use the `cmdstanr` package for interfacing with cmdstan. We will also use the `ggplot2` package for plotting and `dplyr` for data manipulation.

```{r setup, message = FALSE}
library(primarycensored)
library(cmdstanr)
library(ggplot2)
library(dplyr)
```

# Simulating censored and truncated delay distribution data

We'll start by simulating some censored and truncated delay distribution data. We'll use the `rprimarycensored` function (actually we will use the `rpcens ` alias for brevity).

```{r sample-lognormal}
set.seed(123) # For reproducibility

# Define the true distribution parameters
n <- 2000
meanlog <- 1.5
sdlog <- 0.75

# Generate varying pwindow, swindow, and obs_time lengths
pwindows <- sample(1:2, n, replace = TRUE)
swindows <- sample(1:2, n, replace = TRUE)
obs_times <- sample(8:10, n, replace = TRUE)

# Function to generate a single sample
generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rlnorm,
    meanlog = meanlog, sdlog = sdlog,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

# Generate samples
samples <- mapply(generate_sample, pwindows, swindows, obs_times)

# Create initial data frame
delay_data <- data.frame(
  pwindow = pwindows,
  obs_time = obs_times,
  observed_delay = samples,
  observed_delay_upper = samples + swindows
) |>
  mutate(
    observed_delay_upper = pmin(obs_time, observed_delay_upper)
  )

head(delay_data)

# Aggregate to unique combinations and count occurrences
# Aggregate to unique combinations and count occurrences
delay_counts <- delay_data |>
  summarise(
    n = n(),
    .by = c(pwindow, obs_time, observed_delay, observed_delay_upper)
  )

head(delay_counts)

# Compare the samples with and without secondary censoring to the true
# distribution
# Calculate empirical CDF
empirical_cdf <- ecdf(samples)

# Create a sequence of x values for the theoretical CDF
x_seq <- seq(min(samples), max(samples), length.out = 100)

# Calculate theoretical CDF
theoretical_cdf <- plnorm(x_seq, meanlog = meanlog, sdlog = sdlog)

# Create a long format data frame for plotting
cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(c("Observed", "Theoretical"), each = length(x_seq)),
  stringsAsFactors = FALSE
)

# Plot
ggplot(cdf_data, aes(x = x, y = probability, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c(Observed = "#4292C6", Theoretical = "#252525")
  ) +
  geom_vline(
    aes(xintercept = mean(samples), color = "Observed"),
    linetype = "dashed", linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = exp(meanlog + sdlog^2 / 2), color = "Theoretical"),
    linetype = "dashed", linewidth = 1
  ) +
  labs(
    title = "Comparison of Observed vs Theoretical CDF",
    x = "Delay",
    y = "Cumulative Probability",
    color = "CDF Type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

We've aggregated the data to unique combinations of `pwindow`, `swindow`, and `obs_time` and counted the number of occurrences of each `observed_delay` for each combination. This is the data we will use to fit our model.

# Fitting a naive model using cmdstan

We'll start by fitting a naive model using cmdstan. We'll use the `cmdstanr` package to interface with cmdstan. We define the model in a string and then write it to a file as in the [How to use primarycensored with Stan](using-stan-tools.html) vignette.

```{r naive-model}
writeLines(
  "data {
    int<lower=0> N;  // number of observations
    vector[N] y;     // observed delays
    vector[N] n;     // number of occurrences for each delay
  }
  parameters {
    real mu;
    real<lower=0> sigma;
  }
  model {
    // Priors
    mu ~ normal(1, 1);
    sigma ~ normal(0.5, 1);

    // Likelihood
    target += n .* lognormal_lpdf(y | mu, sigma);
  }",
  con = file.path(tempdir(), "naive_lognormal.stan")
)
```

Now let's compile the model

```{r compile-naive-model}
naive_model <- cmdstan_model(file.path(tempdir(), "naive_lognormal.stan"))
```

and now let's fit the compiled model.

```{r fit-naive-model, message = FALSE}
naive_fit <- naive_model$sample(
  data = list(
    # Add a small constant to avoid log(0)
    y = delay_counts$observed_delay + 1e-6,
    n = delay_counts$n,
    N = nrow(delay_counts)
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
naive_fit
```

We see that the model has converged and the diagnostics look good. However, just from the model posterior summary we see that we might not be very happy with the fit. `mu` is smaller than the target `r meanlog` and `sigma` is larger than the target `r sdlog`.

# Fitting an improved model using primarycensored

We'll now fit an improved model using the `primarycensored` package. The main improvement is that we will use the `primarycensored_lpdf` function to fit the model. This is the Stan version of the `pcens()` function and accounts for the primary and secondary censoring windows as well as the truncation. We encode that our primary distribution is a lognormal distribution by passing 1 as the `dist_id` parameter and that our primary event distribution is uniform by passing 1 as the `primary_id` parameter. See the [Stan documentation](https://primarycensored.epinowcast.org/stan/primary__censored__dist_8stan.html#acc97240dee1bc19e4f02013118f3857d) for more details on the `primarycensored_lpdf` function.

```{r pimarycensoreddist-model}
writeLines(
  "
  functions {
    #include primarycensored.stan
    // These functions are required for the primarycensored_lpdf function
    #include primarycensored_ode.stan
    #include primarycensored_analytical_cdf.stan
    #include expgrowth.stan
  }
  data {
    int<lower=0> N;  // number of observations
    array[N] int<lower=0> y;     // observed delays
    array[N] int<lower=0> y_upper;     // observed delays upper bound
    array[N] int<lower=0> n;     // number of occurrences for each delay
    array[N] int<lower=0> pwindow; // primary censoring window
    array[N] int<lower=0> D; // maximum delay
  }
  transformed data {
    array[0] real primary_params;
  }
  parameters {
    real mu;
    real<lower=0> sigma;
  }
  model {
    // Priors
    mu ~ normal(1, 1);
    sigma ~ normal(0.5, 0.5);

    // Likelihood
    for (i in 1:N) {
      target += n[i] * primarycensored_lpmf(
        y[i] | 1, {mu, sigma},
        pwindow[i], y_upper[i], D[i],
        1, primary_params
      );
    }
  }",
  con = file.path(tempdir(), "primarycensored_lognormal.stan")
)
```

Now let's compile the model

```{r compile-primarycensored-model, message = FALSE}
primarycensored_model <- cmdstan_model(
  file.path(tempdir(), "primarycensored_lognormal.stan"),
  include_paths = pcd_stan_path()
)
```

Now let's fit the compiled model.

```{r fit-primarycensored-model, message = FALSE}
primarycensored_fit <- primarycensored_model$sample(
  data = list(
    y = delay_counts$observed_delay,
    y_upper = delay_counts$observed_delay_upper,
    n = delay_counts$n,
    pwindow = delay_counts$pwindow,
    D = delay_counts$obs_time,
    N = nrow(delay_counts)
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
primarycensored_fit
```

We see that the model has converged and the diagnostics look good. We also see that the posterior means are very near the true parameters and the 90% credible intervals include the true parameters.


# Using `pcd_cmdstan_model()` for a more efficient approach

While the previous approach works well, `primarycensored` provides a more efficient and convenient model which we can compile using `pcd_cmdstan_model()`. This approach not only saves time in model specification but also leverages within chain parallelisation to make best use of your machine's resources. Alongside this we also supply a convenience function `pcd_as_stan_data()` to convert our data into a format that can be used to fit the model and supply priors, bounds, and other settings.

Let's use this function to fit our data:

```{r pcd-cmdstan-model, message = FALSE}
# Compile the model with multithreading support
pcd_model <- pcd_cmdstan_model(cpp_options = list(stan_threads = TRUE))

pcd_data <- pcd_as_stan_data(
  delay_counts,
  delay = "observed_delay",
  delay_upper = "observed_delay_upper",
  relative_obs_time = "obs_time",
  dist_id = 1, # Lognormal distribution
  primary_id = 1, # Uniform primary distribution
  param_bounds = list(lower = c(-Inf, 0), upper = c(Inf, Inf)),
  primary_param_bounds = list(lower = numeric(0), upper = numeric(0)),
  priors = list(location = c(1, 0.5), scale = c(1, 0.5)),
  primary_priors = list(location = numeric(0), scale = numeric(0)),
  use_reduce_sum = TRUE # use within chain parallelisation
)

pcd_fit <- pcd_model$sample(
  data = pcd_data,
  chains = 4,
  parallel_chains = 2, # Run 2 chains in parallel
  threads_per_chain = 2, # Use 2 cores per chain
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)

pcd_fit
```

In this model we have a generic `params` vector that contains the parameters for the delay distribution. In this case these are `mu` and `sigma` from the last example. We also have a `primary_params` vector that contains the parameters for the primary distribution. In this case this is empty as we are using a uniform distribution.

We see again that the model has converged and the diagnostics look good. We also see that the posterior means are very near the true parameters and the 90% credible intervals include the true parameters as with the manually written model. **Note that we have set `parallel_chains = 2` and `threads_per_chain = 2` to demonstrate within chain parallelisation. Usually however you would want to use `parallel_chains = <number of chains>` and then use the remainder of the available cores for within chain parallelisation by setting `threads_per_chain = <remaining cores / number of chains>` to ensure that you make best use of your machine's resources.**
