---
title: "Fitting non-parametric delay distributions"
description: "A guide on how to fit non-parametric delay distributions using primarycensored, with both MLE and Bayesian approaches."
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting non-parametric delay distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What we will do

In this vignette we fit a non-parametric delay distribution to doubly censored data using `primarycensored`.
We simulate data from a known piecewise-constant PMF, fit it with MLE via `fitdistdoublecens()` and with Stan via the package model exposed by `pcd_cmdstan_model()`, then compare the four estimates against the truth.
By the end you should be able to recognise when a non-parametric fit is appropriate and reproduce the workflow on your own data.

## What you might want to read first

This vignette assumes familiarity with `vignette("fitting-dists-with-fitdistrplus")` and `vignette("fitting-dists-with-stan")`.
Mathematical derivations of the censored convolution are in `vignette("why-it-works")` and `vignette("analytic-solutions")`.

## Packages used in this vignette

Alongside `primarycensored` we use `cmdstanr` for Stan fitting, `ggplot2` for plotting, and `dplyr` for data manipulation.
MLE fitting goes through `fitdistdoublecens()` (which calls `fitdistrplus`).

```{r setup, message = FALSE}
library(primarycensored)
library(cmdstanr) # nolint: missing_package_linter, object_usage_linter.
library(ggplot2) # nolint: object_usage_linter.
library(dplyr) # nolint: object_usage_linter.
```

# Mathematical background

The non-parametric delay is supported on a fixed grid of $K$ bins with boundaries $b_0 < b_1 < \dots < b_K$.
Two parameterisations of the same step PMF are available.

## Discrete step (direct PMF)

We place a probability $f_i$ on each bin and require $\sum_{i=1}^{K} f_i = 1$ with $f_i \geq 0$.
The CDF is the cumulative sum
$$
F(t) = \sum_{i: b_i \leq t} f_i, \qquad t \in [b_0, b_K].
$$
Under right truncation at $D$ and a censoring window $[d, d_\text{upper}]$ the censored likelihood contribution for one observation is
$$
\frac{F_\text{obs}(d_\text{upper}) - F_\text{obs}(d)}{F_\text{obs}(D)},
$$
where $F_\text{obs}$ is the convolution of $F$ with the primary event density (see `vignette("why-it-works")` and `vignette("analytic-solutions")` for derivations).
The natural prior is a Dirichlet on the simplex, $f \sim \text{Dirichlet}(\alpha)$.

## Discrete hazard (logit-RW)

We instead model the discrete-time hazard $h_i = \Pr(T \in [b_{i-1}, b_i) \mid T \geq b_{i-1})$ on the logit scale via a Gaussian random walk.
With innovations $\varepsilon_j \sim \mathcal{N}(0, 1)$,
$$
\text{logit}(h_i) = \alpha + \sigma \sum_{j=1}^{i-1} \varepsilon_j, \qquad i = 1, \dots, K-1,
$$
and $h_K = 1$ to make the PMF proper.
The PMF is recovered by
$$
f_i = h_i \prod_{j < i} (1 - h_j).
$$
The random walk regularises towards smooth hazards and is well behaved when bins are sparse.
An alternative parameterisation drops the cumulative sum and instead places IID logit random effects around the intercept, $\text{logit}(h_i) = \alpha + \sigma\, \varepsilon_i$, giving independent draws on each bin (`distr = "discretehazardre"`).

# Simulating censored data from a step distribution

We build a bimodal true PMF as a mixture of two discretised lognormals, one at days 2--3 and one at days 6--7.
A shape like this (for example a primary infection peak followed by a secondary reporting bump) is awkward to capture with a single parametric family and is a natural use case for the discrete step distribution.
Here a parametric mixture (e.g. mixture of two lognormals) would also work; this vignette shows the more general non-parametric approach.

```{r truth}
set.seed(123)

K <- 10
boundaries <- 0:K

mode1 <- diff(plnorm(boundaries, meanlog = 0.7, sdlog = 0.4))
mode2 <- diff(plnorm(boundaries, meanlog = 1.9, sdlog = 0.2))
raw_pmf <- 0.7 * mode1 + 0.3 * mode2
true_pmf <- raw_pmf / sum(raw_pmf)
```

Before simulating we plot the two component PMFs together with the resulting bimodal mixture, which makes the non-parametric motivation visible.

```{r truth-plot, fig.width = 7, fig.height = 4}
truth_components <- data.frame(
  bin = rep(seq_len(K), 3),
  pmf = c(
    0.7 * mode1 / sum(raw_pmf),
    0.3 * mode2 / sum(raw_pmf),
    true_pmf
  ),
  series = factor(
    rep(c("component 1", "component 2", "mixture"),
      each = K
    ),
    levels = c("component 1", "component 2", "mixture")
  )
)

ggplot(truth_components, aes(x = bin, y = pmf, fill = series)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Bin", y = "Probability", fill = NULL,
    title = "True PMF and its lognormal components"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )
```

We now simulate `n = 500` doubly censored observations using `rpcens()`.
We let the right-truncation point vary across observations, so a meaningful fraction of draws are right-truncated below the maximum support; the fits must therefore correctly account for per-observation truncation as well as for primary and secondary censoring.

```{r simulate}
n <- 500
pwindows <- rep(1, n)
swindows <- rep(1, n)
obs_times <- sample(8:11, n, replace = TRUE)

generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rdiscretestep,
    boundaries = boundaries, pmf = true_pmf,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

samples <- mapply(generate_sample, pwindows, swindows, obs_times)

delay_data <- data.frame(
  delay       = samples,
  delay_upper = samples + swindows,
  pwindow     = pwindows,
  D           = obs_times
)

head(delay_data)
```

We can compare the empirical CDF of the simulated data against the true step CDF.

```{r plot-simulation, fig.width = 7, fig.height = 4}
empirical_cdf <- ecdf(samples)
x_seq <- seq(0, K, length.out = 200)
theoretical_cdf <- pdiscretestep(
  x_seq,
  boundaries = boundaries, pmf = true_pmf
)

cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(c("Observed", "Theoretical"), each = length(x_seq))
)

ggplot(cdf_data, aes(x = x, y = probability, colour = type)) +
  geom_step(linewidth = 1) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    x = "Delay", y = "Cumulative probability",
    colour = NULL,
    title = "Observed vs theoretical CDF"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  coord_cartesian(xlim = c(0, K))
```

The observed CDF is shifted left relative to the truth because right truncation and primary censoring suppress long delays in the raw data.
Our goal is to recover the true PMF from these observed values.

# Fitting via `fitdistdoublecens()`

`fitdistdoublecens()` accepts two non-parametric options: `distr = "discretestep"` and `distr = "discretehazard"`.
We start with the hazard parameterisation because, in our experience, it is more robust at the optimisation stage and faster to converge; we then show the direct PMF fit and discuss why it sometimes complains.

## Discrete hazard fit

```{r fit-haz, message = FALSE}
haz_start <- c(
  list(alpha = -2, log_sigma = log(1.0)),
  as.list(stats::setNames(
    rep(0, K - 1L), paste0("eps_", seq_len(K - 1L))
  ))
)

fit_haz <- fitdistdoublecens(
  delay_data,
  distr      = "discretehazard",
  start      = haz_start,
  left       = "delay",
  right      = "delay_upper",
  pwindow    = "pwindow",
  D          = "D",
  boundaries = boundaries,
  control    = list(maxit = 10000)
)

summary(fit_haz)
```

We also fit the IID logit random-effect variant on the same data to compare it against the random walk later on.
The RE model drops the smoothing between adjacent bins, so its hazards are independent draws around the intercept rather than a smooth trajectory.
With `K = 10` bins on this bimodal dataset the RE likelihood is flat in `log_sigma` at the eps-zero start (every bin equals the intercept), so optimisation through `fitdistrplus` may not converge from this initialisation; we wrap the call in `tryCatch()` and skip the RE estimate in the comparison plot if so.

```{r fit-haz-re, message = FALSE}
haz_re_start <- c(
  list(alpha = -2, log_sigma = log(0.5)),
  as.list(stats::setNames(
    rep(0.1 * seq_len(K - 1L), 1L), paste0("eps_", seq_len(K - 1L))
  ))
)

fit_haz_re <- tryCatch(
  fitdistdoublecens(
    delay_data,
    distr      = "discretehazardre",
    start      = haz_re_start,
    left       = "delay",
    right      = "delay_upper",
    pwindow    = "pwindow",
    D          = "D",
    boundaries = boundaries,
    calcvcov   = FALSE,
    control    = list(maxit = 50000)
  ),
  error = function(e) {
    message("Could not fit RE variant: ", conditionMessage(e))
    NULL
  }
)
```

## Discrete step fit

```{r fit-step, message = FALSE}
step_start <- as.list(stats::setNames(
  rep(1 / K, K - 1L), paste0("p", seq_len(K - 1L))
))

fit_step <- fitdistdoublecens(
  delay_data,
  distr      = "discretestep",
  start      = step_start,
  left       = "delay",
  right      = "delay_upper",
  pwindow    = "pwindow",
  D          = "D",
  boundaries = boundaries
)

summary(fit_step)
```

The direct PMF fit often emits a `cov2cor` warning at the `summary()` step.
This comes from `fitdistrplus` taking a square root of negative or zero diagonal entries of the variance--covariance matrix.
With this much primary censoring and right truncation the last few bins (whose probability mass is small and partially smeared by truncation) sit close to the boundary of the simplex, so the Hessian is near singular.
The point estimates remain usable, but standard errors for those bins are not trustworthy.
The hazard fit avoids this because it parameterises the simplex through unconstrained logit-hazards.

# Fitting via Stan

The package model returned by `pcd_cmdstan_model()` accepts a `nonparametric` block that selects either `dist_id = 26` (Dirichlet on the PMF) or `dist_id = 27` (random-walk hazards) directly, with `boundaries` carried as data and the simplex / random-walk constructed inside the Stan model.
We use the same model object for both fits below; only the `nonparametric` argument to `pcd_as_stan_data()` changes.

The current analytic step CDF in Stan returns exactly zero for delays at the lower support boundary plus one (here `delay = 1`), which makes the autodiff gradient non-finite during HMC.
We therefore filter `delay > 1` for the Stan fit; this is a workaround and the affected mass is small.
Fixing the underlying analytic convolution (so the right-continuous step semantics are reflected in every interval contribution, not only at the upper boundary) is tracked as follow-up work.
We aggregate to unique `(pwindow, D, delay, delay_upper)` rows for efficiency.

```{r aggregate}
delay_counts <- delay_data |>
  filter(delay > 1) |>
  summarise(
    n = n(),
    .by = c(pwindow, D, delay, delay_upper)
  ) |>
  rename(relative_obs_time = D)
```

```{r np-model}
np_model <- pcd_cmdstan_model()
```

## Dirichlet prior on the PMF

We put a Dirichlet prior on the PMF.
The package model declares the simplex parameter as `simplex[K] np_pmf` directly when `paramtype = "simplex"`, which avoids any normalisation trick.

```{r dirichlet-model-run, message = FALSE}
dirichlet_data <- pcd_as_stan_data(
  delay_counts,
  dist_id = 26,
  primary_id = 1,
  param_bounds = list(lower = numeric(0), upper = numeric(0)),
  primary_param_bounds = list(lower = numeric(0), upper = numeric(0)),
  priors = list(location = numeric(0), scale = numeric(0)),
  primary_priors = list(location = numeric(0), scale = numeric(0)),
  nonparametric = list(
    K          = K,
    boundaries = boundaries,
    paramtype  = "simplex"
  )
)

dirichlet_fit <- np_model$sample(
  data            = dirichlet_data,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 500,
  iter_sampling   = 500,
  refresh         = ifelse(interactive(), 50, 0),
  show_messages   = interactive()
)

dirichlet_fit$summary(
  variables = paste0("np_pmf[", seq_len(K), "]")
)
```

## Gaussian random walk on the logit hazards

The package model also exposes the hazard random-walk parameterisation when `paramtype = "hazard"`.
The final bin hazard is pinned to 1 inside the model so the PMF sums to 1; only the prior hyperparameters are user-supplied.

```{r rw-model-run, message = FALSE}
rw_data <- pcd_as_stan_data(
  delay_counts,
  dist_id = 27,
  primary_id = 1,
  param_bounds = list(lower = numeric(0), upper = numeric(0)),
  primary_param_bounds = list(lower = numeric(0), upper = numeric(0)),
  priors = list(location = numeric(0), scale = numeric(0)),
  primary_priors = list(location = numeric(0), scale = numeric(0)),
  nonparametric = list(
    K = K,
    boundaries = boundaries,
    paramtype = "hazard",
    hazard_priors = list(
      alpha_mean     = 0, alpha_sd     = 5,
      log_sigma_mean = 0, log_sigma_sd = 1
    )
  )
)

rw_fit <- np_model$sample(
  data            = rw_data,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 500,
  iter_sampling   = 500,
  refresh         = ifelse(interactive(), 50, 0),
  show_messages   = interactive()
)

rw_fit$summary(
  variables = paste0("np_weights[", seq_len(K), "]")
)
```

# Comparing estimates against the truth

We extract the four sets of PMF estimates and plot them on a single panel.
We use shape to encode the parameterisation (step vs hazard) and colour to encode the method (MLE vs Stan).
Stan estimates are shown with `geom_pointrange` (median plus 90% credible interval); MLE estimates are shown with `geom_point`.

```{r extract-mle}
extract_step_pmf <- function(fit, K) {
  p_free <- coef(fit)[paste0("p", seq_len(K - 1))]
  c(p_free, 1 - sum(p_free))
}

extract_haz_pmf <- function(fit, K, model = c("rw", "re")) {
  model <- match.arg(model)
  alpha_hat <- coef(fit)[["alpha"]]
  log_sigma_hat <- coef(fit)[["log_sigma"]]
  sigma_hat <- exp(log_sigma_hat)
  eps_hat <- coef(fit)[paste0("eps_", seq_len(K - 1))]
  deltas <- if (model == "rw") cumsum(c(0, eps_hat)) else c(0, eps_hat)
  logit_h <- alpha_hat + sigma_hat * deltas
  h <- 1 / (1 + exp(-logit_h))
  h[K] <- 1
  hazards_to_pmf(h)
}

pmf_step <- extract_step_pmf(fit_step, K)
pmf_haz <- extract_haz_pmf(fit_haz, K, "rw")
pmf_haz_re <- if (!is.null(fit_haz_re)) {
  extract_haz_pmf(fit_haz_re, K, "re")
} else {
  NULL
}
```

```{r extract-stan-intervals}
dirichlet_draws <- dirichlet_fit$draws(
  variables = paste0("np_pmf[", seq_len(K), "]"),
  format    = "matrix"
)

dirichlet_summary <- data.frame(
  bin    = seq_len(K),
  median = apply(dirichlet_draws, 2, median),
  lo90   = apply(dirichlet_draws, 2, quantile, 0.05),
  hi90   = apply(dirichlet_draws, 2, quantile, 0.95)
)

hazard_draws <- rw_fit$draws(
  variables = paste0("np_weights[", seq_len(K), "]"),
  format    = "matrix"
)

pmf_draws_rw <- t(apply(hazard_draws, 1, function(h) {
  h[K] <- 1
  hazards_to_pmf(h)
}))

rw_summary <- data.frame(
  bin    = seq_len(K),
  median = apply(pmf_draws_rw, 2, median),
  lo90   = apply(pmf_draws_rw, 2, quantile, 0.05),
  hi90   = apply(pmf_draws_rw, 2, quantile, 0.95)
)
```

```{r comparison-plot, fig.width = 8, fig.height = 5}
mle_df <- rbind(
  data.frame(
    bin = seq_len(K), pmf = pmf_step,
    param = "step", method = "MLE"
  ),
  data.frame(
    bin = seq_len(K), pmf = pmf_haz,
    param = "hazard (RW)", method = "MLE"
  )
)
if (!is.null(pmf_haz_re)) {
  mle_df <- rbind(
    mle_df,
    data.frame(
      bin = seq_len(K), pmf = pmf_haz_re,
      param = "hazard (RE)", method = "MLE"
    )
  )
}

stan_df <- rbind(
  transform(dirichlet_summary, param = "step", method = "Stan"),
  transform(rw_summary, param = "hazard (RW)", method = "Stan")
)

truth_df <- data.frame(bin = seq_len(K), pmf = true_pmf)

dodge <- position_dodge(width = 0.6)

ggplot() +
  geom_col(
    data = truth_df, aes(x = bin, y = pmf),
    fill = "#D9D9D9", colour = "#252525", width = 0.6
  ) +
  geom_pointrange(
    data = stan_df,
    aes(
      x = bin, y = median, ymin = lo90, ymax = hi90,
      colour = method, shape = param
    ),
    position = dodge, size = 0.5
  ) +
  geom_point(
    data = mle_df,
    aes(x = bin, y = pmf, colour = method, shape = param),
    position = dodge, size = 2.5
  ) +
  scale_colour_brewer(palette = "Set2") +
  scale_shape_manual(
    values = c(step = 16, `hazard (RW)` = 17, `hazard (RE)` = 15)
  ) +
  labs(
    x = "Bin", y = "Probability",
    colour = "Method", shape = "Parameterisation",
    title = "PMF estimates vs truth"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

All four methods recover the bimodal shape of the truth.
Bin 1 is biased low for the Stan fits because we filtered the `delay = 1` rows; with the analytic step CDF fix in place this would not be necessary.
Bins 3 and 4 sit between the two modes and absorb most of the remaining disagreement: under primary censoring with `pwindow = 1` the mass that the convolution puts at days 2--3 leaks into days 3--4, which weakens identifiability for those middle bins.
The Stan posterior intervals correctly widen there to reflect that the data are less informative about the trough than about the peaks.
With more observations, narrower primary windows, or fewer bins around the trough, the disagreement between methods on those bins shrinks.

# When to choose each parameterisation

The hazard / random-walk parameterisation is smoother and more robust at both MLE and Stan stages, so we suggest it as the default unless you have a reason to expect sharp jumps in the PMF.
The Dirichlet on the PMF is more flexible but can be poorly identified for bins where the censored convolution has little leverage, which is exactly where you saw the `cov2cor` warning.
This area is active research; for a recent overview see Charniga et al. [@Charniga2024].

In many applied settings a parametric distribution is a better fit than a non-parametric one, both because it has fewer parameters and because the smoothness it assumes matches what we know about biological delays.
For parametric alternatives see `vignette("fitting-dists-with-fitdistrplus")` and `vignette("fitting-dists-with-stan")`.
For more flexible delay distribution fitting see the [`epidist`](https://epidist.epinowcast.org) package.

## How you might adapt this vignette

- Adjust `K` and `boundaries` to match the temporal resolution of your data.
- Pass a custom non-uniform `boundaries` vector to capture daily, weekly, or irregular intervals.
- Swap `delay_data` for your own doubly censored observations.
- Change the Dirichlet concentration `alpha` to express stronger or weaker prior beliefs.
- Adjust `sigma ~ normal(0, 1)` to allow more or less smoothness in the logit-hazard trajectory.
- Use any registered primary distribution (uniform or expgrowth here) by passing the matching `primary_id` and `primary_params`.
