---
title: "Fitting non-parametric delay distributions"
description: "A guide on how to fit non-parametric delay distributions using primarycensored, with both MLE and Bayesian approaches."
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting non-parametric delay distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What are we going to do in this vignette

In this vignette, we demonstrate how to fit a non-parametric delay
distribution to doubly censored data using `primarycensored`.
We cover the following key points:

1. Simulating censored data from a known piecewise-constant distribution
2. Fitting via `fitdistdoublecens()` using both the direct PMF
   (`distr = "discretestep"`) and logit-hazard random-walk
   (`distr = "discretehazard"`) parameterisations
3. Fitting via Stan using a Dirichlet-on-PMF prior and a
   random-walk-on-logit-hazard prior
4. Comparing all four estimates against the true PMF, including
   posterior uncertainty from the Bayesian fits
5. Demonstrating the analytic primary convolution with an exponential-growth
   primary distribution

## What you will learn

By the end of this vignette, you will be able to:

- Use `rdiscretestep()`, `pdiscretestep()`, and `ddiscretestep()` to work
  with piecewise-constant distributions
- Fit a non-parametric step distribution to doubly censored data using
  `fitdistdoublecens()` with `distr = "discretestep"` or
  `distr = "discretehazard"`
- Write custom Stan models that call `primarycensored_lpmf` with
  `dist_id = 26` (direct PMF) or `dist_id = 27` (logit-hazard)
- Use the analytic primary convolution with non-uniform primary distributions

## What might I need to know before starting

This vignette assumes familiarity with the concepts in
`vignette("fitting-dists-with-fitdistrplus")` and
`vignette("fitting-dists-with-stan")`.
Non-parametric distributions are useful when you do not want to assume a
particular functional form for the delay distribution.

## Packages used in this vignette

Alongside the `primarycensored` package we will use `cmdstanr` for Stan
fitting, `ggplot2` for plotting, and `dplyr` for data manipulation. MLE
fitting is delegated to `fitdistrplus` via `fitdistdoublecens()`.

```{r setup, message = FALSE}
library(primarycensored)
library(cmdstanr) # nolint: missing_package_linter, object_usage_linter.
library(ggplot2) # nolint: object_usage_linter.
library(dplyr) # nolint: object_usage_linter.
```

# Simulating censored data from a step distribution

We simulate data from a known 14-bin step distribution using `rpcens()`.
The true PMF is obtained by discretising a lognormal(meanlog = 1.5,
sdlog = 0.6) distribution onto integer bins 1–14, giving a distribution
that peaks around days 3–4 and has a right tail, as one might expect for
an incubation or reporting delay.
We simulate under right-truncation at `D = K + 1 = 15`, so the fits must
account for this.

```{r simulate}
set.seed(123) # For reproducibility

n <- 500
K <- 14
boundaries <- 0:K
D <- K + 1 # right-truncation point

# Discretise lognormal onto bins 1..K then normalise
raw_pmf <- diff(plnorm(boundaries, meanlog = 1.5, sdlog = 0.6))
true_pmf <- raw_pmf / sum(raw_pmf)

pwindows <- rep(1, n)
swindows <- rep(1, n)
obs_times <- rep(D, n)

generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rdiscretestep,
    boundaries = boundaries, pmf = true_pmf,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

samples <- mapply(generate_sample, pwindows, swindows, obs_times)

delay_data <- data.frame(
  delay       = samples,
  delay_upper = samples + swindows,
  pwindow     = pwindows,
  D           = obs_times
)

head(delay_data)
```

`rprimarycensored()` uses rejection sampling.
It generates primary event times and delays, computes the total delay,
and discards any sample with `total_delay >= D`, re-sampling until `n`
valid observations are obtained.
This correctly represents the truncation mechanism rather than clipping
observed values.

We can compare the empirical CDF of the simulated data against the true
step CDF.

```{r plot-simulation}
empirical_cdf <- ecdf(samples)
x_seq <- seq(0, K, length.out = 200)
theoretical_cdf <- pdiscretestep(
  x_seq,
  boundaries = boundaries, pmf = true_pmf
)

cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(
    c("Observed", "Theoretical"),
    each = length(x_seq)
  ),
  stringsAsFactors = FALSE
)

ggplot(cdf_data, aes(x = x, y = probability, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c(Observed = "#4292C6", Theoretical = "#252525")
  ) +
  labs(
    title = "Observed vs Theoretical CDF (step distribution)",
    x     = "Delay",
    y     = "Cumulative probability",
    color = "CDF type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor  = element_blank(),
    plot.title        = element_text(hjust = 0.5),
    legend.position   = "bottom"
  ) +
  coord_cartesian(xlim = c(0, K))
```

The observed CDF is shifted left relative to the truth because right
truncation and primary censoring suppress long delays in the raw data.
Our goal is to recover the true PMF from these observed values.

# Fitting via `fitdistdoublecens()`

`fitdistdoublecens()` accepts two non-parametric options:
`distr = "discretestep"` and `distr = "discretehazard"`.
The number of bins is inferred from `length(start)`, and both share the
column naming convention of the parametric interface.

## Direct PMF parameterisation (`distr = "discretestep"`)

The `"discretestep"` parameterisation estimates `K - 1` free bin
probabilities directly.
The last bin probability is set to `1 - sum(p1, ..., p_{K-1})`, which
keeps the PMF normalised.

```{r fit-step, message = FALSE}
step_start <- as.list(stats::setNames(
  rep(1 / K, K - 1L), paste0("p", seq_len(K - 1L))
))

fit_step <- fitdistdoublecens(
  delay_data,
  distr      = "discretestep",
  start      = step_start,
  left       = "delay",
  right      = "delay_upper",
  pwindow    = "pwindow",
  D          = "D",
  boundaries = boundaries,
  control    = list(maxit = 20000)
)

summary(fit_step)
```

The estimated parameters `p1` through `p13` correspond to the first 13
bin probabilities; the 14th is determined by the sum-to-one constraint.

## Logit-hazard random-walk parameterisation (`distr = "discretehazard"`)

The `"discretehazard"` parameterisation models discrete-time hazards on
the logit scale via a Gaussian random walk, which smooths the estimated
PMF and regularises against over-fitting to noise.
A MAP-style prior penalty (equivalent to
`alpha ~ normal(0, 5)`, `log_sigma ~ normal(log(0.5), 1)`,
`eps_j ~ normal(0, 1)`) is applied.

```{r fit-haz, message = FALSE}
haz_start <- c(
  list(alpha = -2, log_sigma = log(0.3)),
  as.list(stats::setNames(
    rep(0, K - 1L), paste0("eps_", seq_len(K - 1L))
  ))
)

fit_haz <- fitdistdoublecens(
  delay_data,
  distr      = "discretehazard",
  start      = haz_start,
  left       = "delay",
  right      = "delay_upper",
  pwindow    = "pwindow",
  D          = "D",
  boundaries = boundaries,
  control    = list(maxit = 50000, reltol = 1e-8)
)

summary(fit_haz)
```

# Fitting via Stan

We show two model variants using the non-parametric Stan distributions.
`dist_id = 26` is the direct-PMF (discretestep) parameterisation;
its parameter vector packs `K + 1` boundaries followed by `K` PMF values,
so `n_params = 2 * K + 1`.
`dist_id = 27` is the logit-hazard (discretehazard) parameterisation;
its parameter vector also packs `K + 1` boundaries followed by `K` hazard
values, where the last hazard is constrained to 1.

## Dirichlet-on-PMF model

This model places a symmetric Dirichlet prior on the PMF by drawing
`K` independent `gamma(alpha, 1)` variates and normalising.
Passing the packed `[boundaries, pmf]` vector as `params` in
`primarycensored_lpmf` dispatches to the analytic step-distribution
integration in the package.

```{r dirichlet-model-run, message = FALSE}
dirichlet_stan_code <- "
functions {
  #include primarycensored.stan
  #include primarycensored_ode.stan
  #include primarycensored_analytical_cdf.stan
  #include expgrowth.stan
  #include nonparametric.stan
}
data {
  int<lower=1> N;
  array[N] int<lower=0> y;
  array[N] int<lower=0> y_upper;
  array[N] int<lower=0> n_obs;
  array[N] int<lower=0> pwindow;
  array[N] int<lower=0> D;
  int<lower=1> K;
  vector<lower=0>[K+1] boundaries;
  vector<lower=0>[K] alpha;
}
transformed data {
  array[0] real primary_params;
}
parameters {
  vector<lower=0>[K] gamma_raw;
}
transformed parameters {
  vector[K] pmf = gamma_raw / sum(gamma_raw);
  array[2*K+1] real params;
  for (k in 1:(K+1)) params[k]     = boundaries[k];
  for (k in 1:K)     params[K+1+k] = pmf[k];
}
model {
  gamma_raw ~ gamma(alpha, 1);
  for (i in 1:N) {
    target += n_obs[i] * primarycensored_lpmf(
      y[i] | 26, params,
      pwindow[i], y_upper[i], 0, D[i],
      1, primary_params
    );
  }
}
"

dirichlet_stan_file <- file.path(
  tempdir(), "step_dirichlet.stan"
)
writeLines(dirichlet_stan_code, dirichlet_stan_file)

dirichlet_model <- cmdstan_model(
  dirichlet_stan_file,
  include_paths = pcd_stan_path()
)

delay_counts <- delay_data |>
  summarise(
    n_obs = n(),
    .by   = c(pwindow, D, delay, delay_upper)
  ) |>
  rename(y = delay, y_upper = delay_upper)

dirichlet_data <- list(
  N          = nrow(delay_counts),
  y          = delay_counts$y,
  y_upper    = delay_counts$y_upper,
  n_obs      = delay_counts$n_obs,
  pwindow    = delay_counts$pwindow,
  D          = delay_counts$D,
  K          = K,
  boundaries = as.numeric(boundaries),
  alpha      = rep(1, K) # symmetric Dirichlet(1, ..., 1)
)

dirichlet_fit <- dirichlet_model$sample(
  data            = dirichlet_data,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 500,
  iter_sampling   = 500,
  refresh         = ifelse(interactive(), 50, 0),
  show_messages   = interactive()
)

dirichlet_fit$summary(
  variables = paste0("pmf[", seq_len(K), "]")
)
```

## Random-walk-on-logit-hazard model

This model places a Gaussian random walk prior on the logit-hazard at
each time point.
The final bin hazard is constrained to 1, which ensures the PMF sums to 1.

In the Stan code the `transformed parameters` block builds a `vector[K]
hazards` (with `hazards[K] = 1`) and packs the parameter array as
`[boundaries (K+1), hazards (K)]`.
This is passed to `primarycensored_lpmf` with `dist_id = 27`, which
dispatches directly to the hazard-parameterised analytic convolution.
No manual PMF conversion is needed.

The Stan model uses a half-Normal prior on `sigma` for simplicity; the MLE
penalty is on `log_sigma` and is therefore not identical, though both impose
comparable smoothness.

```{r rw-model-run, message = FALSE}
rw_stan_code <- "
functions {
  #include primarycensored.stan
  #include primarycensored_ode.stan
  #include primarycensored_analytical_cdf.stan
  #include expgrowth.stan
  #include nonparametric.stan
}
data {
  int<lower=1> N;
  array[N] int<lower=0> y;
  array[N] int<lower=0> y_upper;
  array[N] int<lower=0> n_obs;
  array[N] int<lower=0> pwindow;
  array[N] int<lower=0> D;
  int<lower=1> K;
  vector<lower=0>[K+1] boundaries;
}
transformed data {
  array[0] real primary_params;
}
parameters {
  real alpha;
  real<lower=0> sigma;
  vector[K-1] eps;
}
transformed parameters {
  vector[K] logit_h;
  logit_h[1] = alpha;
  for (i in 2:K) logit_h[i] = logit_h[i-1] + sigma * eps[i-1];
  vector[K] hazards = inv_logit(logit_h);
  hazards[K] = 1;
  array[2*K+1] real params;
  for (k in 1:(K+1)) params[k]     = boundaries[k];
  for (k in 1:K)     params[K+1+k] = hazards[k];
}
model {
  alpha ~ normal(0, 5);
  sigma ~ normal(0, 1);
  eps   ~ std_normal();
  for (i in 1:N) {
    target += n_obs[i] * primarycensored_lpmf(
      y[i] | 27, params,
      pwindow[i], y_upper[i], 0, D[i],
      1, primary_params
    );
  }
}
"

rw_stan_file <- file.path(tempdir(), "step_rw.stan")
writeLines(rw_stan_code, rw_stan_file)

rw_model <- cmdstan_model(
  rw_stan_file,
  include_paths = pcd_stan_path()
)

rw_data <- list(
  N          = nrow(delay_counts),
  y          = delay_counts$y,
  y_upper    = delay_counts$y_upper,
  n_obs      = delay_counts$n_obs,
  pwindow    = delay_counts$pwindow,
  D          = delay_counts$D,
  K          = K,
  boundaries = as.numeric(boundaries)
)

rw_fit <- rw_model$sample(
  data            = rw_data,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 500,
  iter_sampling   = 500,
  refresh         = ifelse(interactive(), 50, 0),
  show_messages   = interactive()
)

rw_fit$summary(
  variables = paste0("hazards[", seq_len(K), "]")
)
```

# Comparing estimates against the truth

We extract all four sets of estimates and show them in a single plot.
The truth is shown as a bar chart; MLE point estimates are shown as
coloured points; Bayesian posterior summaries are shown as points with
90% credible intervals derived from the posterior draws.

```{r extract-mle}
# Helper: extract PMF from discretestep MLE fit
extract_step_pmf <- function(fit, K) {
  p_free <- coef(fit)[paste0("p", seq_len(K - 1))]
  c(p_free, 1 - sum(p_free))
}

# Helper: reconstruct PMF from discretehazard MLE fit
extract_haz_pmf <- function(fit, K) {
  alpha_hat <- coef(fit)[["alpha"]]
  log_sigma_hat <- coef(fit)[["log_sigma"]]
  sigma_hat <- exp(log_sigma_hat)
  eps_hat <- coef(fit)[paste0("eps_", seq_len(K - 1))]
  deltas <- c(0, eps_hat)
  logit_h <- alpha_hat + sigma_hat * cumsum(deltas)
  h <- 1 / (1 + exp(-logit_h))
  h[K] <- 1
  hazards_to_pmf(h)
}

pmf_step <- extract_step_pmf(fit_step, K)
pmf_haz <- extract_haz_pmf(fit_haz, K)
```

```{r extract-stan-intervals}
# Dirichlet model: pmf[k] is directly available as a parameter
dirichlet_draws <- dirichlet_fit$draws(
  variables = paste0("pmf[", seq_len(K), "]"),
  format    = "matrix"
)

dirichlet_summary <- data.frame(
  bin    = seq_len(K),
  median = apply(dirichlet_draws, 2, median),
  lo90   = apply(dirichlet_draws, 2, quantile, 0.05),
  hi90   = apply(dirichlet_draws, 2, quantile, 0.95)
)

# RW-hazard model: compute PMF from each draw of hazards
hazard_draws <- rw_fit$draws(
  variables = paste0("hazards[", seq_len(K), "]"),
  format    = "matrix"
)

# Convert each draw's hazard vector to a PMF
pmf_draws_rw <- t(apply(hazard_draws, 1, function(h) {
  h[K] <- 1
  hazards_to_pmf(h)
}))

rw_summary <- data.frame(
  bin    = seq_len(K),
  median = apply(pmf_draws_rw, 2, median),
  lo90   = apply(pmf_draws_rw, 2, quantile, 0.05),
  hi90   = apply(pmf_draws_rw, 2, quantile, 0.95)
)
```

```{r comparison-plot, fig.width = 8, fig.height = 5}
# Colour palette
pal <- c(
  truth          = "#252525",
  discretestep   = "#4292C6",
  discretehazard = "#E6550D",
  dirichlet_stan = "#31A354",
  rw_stan        = "#756BB1"
)

# MLE point data
mle_df <- rbind(
  data.frame(
    bin = seq_len(K), pmf = pmf_step,
    method = "discretestep", stringsAsFactors = FALSE
  ),
  data.frame(
    bin = seq_len(K), pmf = pmf_haz,
    method = "discretehazard", stringsAsFactors = FALSE
  )
)

# Stan interval data
stan_df <- rbind(
  mutate(dirichlet_summary, method = "dirichlet_stan"),
  mutate(rw_summary, method = "rw_stan")
)

# Truth bars
truth_df <- data.frame(
  bin = seq_len(K), pmf = true_pmf, stringsAsFactors = FALSE
)

ggplot() +
  # Truth as grey bars
  geom_col(
    data = truth_df,
    aes(x = bin, y = pmf),
    fill = "#D9D9D9",
    colour = "#252525",
    width = 0.6
  ) +
  # Stan 90% CI ribbons
  geom_errorbar(
    data = stan_df,
    aes(x = bin, ymin = lo90, ymax = hi90, colour = method),
    width = 0.25,
    linewidth = 0.8,
    position = position_dodge(width = 0.5)
  ) +
  # Stan posterior medians
  geom_point(
    data = stan_df,
    aes(x = bin, y = median, colour = method, shape = method),
    size = 2.5,
    position = position_dodge(width = 0.5)
  ) +
  # MLE points
  geom_point(
    data = mle_df,
    aes(x = bin, y = pmf, colour = method, shape = method),
    size = 2.5,
    position = position_dodge(width = 0.5)
  ) +
  scale_colour_manual(
    values = pal[c(
      "discretestep", "discretehazard",
      "dirichlet_stan", "rw_stan"
    )],
    labels = c(
      discretestep   = "discretestep MLE",
      discretehazard = "discretehazard MLE",
      dirichlet_stan = "Dirichlet Stan (median + 90% CI)",
      rw_stan        = "RW-hazard Stan (median + 90% CI)"
    )
  ) +
  scale_shape_manual(
    values = c(
      discretestep   = 16,
      discretehazard = 17,
      dirichlet_stan = 15,
      rw_stan        = 18
    ),
    labels = c(
      discretestep   = "discretestep MLE",
      discretehazard = "discretehazard MLE",
      dirichlet_stan = "Dirichlet Stan (median + 90% CI)",
      rw_stan        = "RW-hazard Stan (median + 90% CI)"
    )
  ) +
  labs(
    title = "PMF estimates vs truth (bars = truth, points = estimates)",
    x     = "Bin",
    y     = "Probability"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title       = element_text(hjust = 0.5),
    legend.position  = "bottom",
    legend.title     = element_blank()
  ) +
  guides(
    colour = guide_legend(nrow = 2),
    shape  = guide_legend(nrow = 2)
  )
```

All four methods recover the true PMF (grey bars) well.
The Stan 90% credible intervals correctly bracket the truth across most bins,
and the MLE point estimates fall within or very close to those intervals.
Residual discrepancies reflect both sampling variability (n = 500) and the
mild prior differences between parameterisations.

# Analytic convolution with a non-uniform primary

The analytic primary convolution for step distributions handles any
primary distribution with a known CDF, not only the uniform.
Below we demonstrate fitting when primary events follow an exponential-growth
distribution (`dexpgrowth` / `pexpgrowth`) with growth rate `r` = 0.5.

## Simulating with exponential-growth primary

```{r expgrowth-simulate}
set.seed(456)

generate_expgrowth_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rdiscretestep, # nolint: object_usage_linter
    boundaries = boundaries, pmf = true_pmf,
    pwindow = pwindow, swindow = swindow, D = obs_time,
    rprimary = rexpgrowth,
    rprimary_args = list(r = 0.5)
  )
}

samples_eg <- mapply(
  generate_expgrowth_sample, pwindows, swindows, obs_times
)

delay_data_eg <- data.frame(
  delay       = samples_eg,
  delay_upper = samples_eg + swindows,
  pwindow     = pwindows,
  D           = obs_times
)
```

## Fitting with `fitdistdoublecens()` and exponential-growth primary

Passing `dprimary = dexpgrowth` and `primary_args = list(r = 0.5)` to
`fitdistdoublecens()` tells the fitting routine to use the same primary
distribution used to generate the data.
The analytic convolution is dispatched automatically because `pexpgrowth`
is registered alongside `dexpgrowth` in the primary distribution registry.

```{r expgrowth-fit, message = FALSE}
fit_eg <- fitdistdoublecens(
  delay_data_eg,
  distr        = "discretestep",
  start        = step_start,
  left         = "delay",
  right        = "delay_upper",
  pwindow      = "pwindow",
  D            = "D",
  boundaries   = boundaries,
  dprimary     = dexpgrowth,
  primary_args = list(r = 0.5),
  control      = list(maxit = 20000)
)

pmf_eg <- extract_step_pmf(fit_eg, K)

data.frame(
  bin           = seq_len(K),
  truth         = round(true_pmf, 3),
  expgrowth_mle = round(pmf_eg, 3)
)
```

The recovered PMF is close to the truth, demonstrating that the analytic
method correctly accounts for the non-uniform primary distribution.
Using the analytic convolution avoids the numerical errors that `integrate()`
can introduce at the piecewise-constant step boundaries.

# When to choose each parameterisation

The `"discretestep"` / Dirichlet parameterisation places no structural
constraint on the shape of the PMF.
It works well when the number of bins is small and the data are plentiful.
The `"discretehazard"` / random-walk parameterisation imposes smoothness
via the random walk prior, which can be valuable with sparse data or many
bins, at the cost of some flexibility in capturing sharp peaks.
For most applied analyses with fewer than fifteen bins the two give similar
results, so the choice is largely a matter of preference.
When in doubt, the `"discretestep"` approach is simpler to interpret
because its parameters are bin probabilities directly.

## Summary

In this vignette we have shown how to fit a non-parametric step
distribution to doubly censored data using `primarycensored`.
The key takeaways are:

- **Use `fitdistdoublecens()` with `distr = "discretestep"` or
  `distr = "discretehazard"`** for MLE-based non-parametric fitting.
- **Stan `dist_id = 26`** identifies the direct-PMF parameterisation.
  Pack parameters as `[boundaries (K+1), pmf (K)]` in the `params` array.
- **Stan `dist_id = 27`** identifies the logit-hazard parameterisation.
  Pack parameters as `[boundaries (K+1), hazards (K)]`; the analytic
  convolution handles the rest.
- **Dirichlet-on-PMF** and **random-walk-on-logit-hazard** are two natural
  Bayesian priors; the latter encourages smoother distributions.
- The analytic primary convolution supports any primary distribution with
  a known CDF, including exponential growth.
- Properly accounting for primary censoring and truncation is as important
  here as with parametric distributions.

For parametric alternatives, see `vignette("fitting-dists-with-fitdistrplus")`
and `vignette("fitting-dists-with-stan")`.
For more flexible delay distribution fitting, see the
[`epidist`](https://epidist.epinowcast.org) package.

## How you might adapt this vignette

- **Change the number of bins**: Adjust `K` and `boundaries` to match
  the temporal resolution of your data.
- **Use non-uniform boundaries**: Pass a custom `boundaries` vector to
  capture daily, weekly, or irregular intervals.
- **Replace simulation with real data**: Swap out `delay_data` with your
  own doubly censored observations.
- **Adjust the Dirichlet concentration**: Change `alpha` to express
  stronger or weaker prior beliefs about bin probabilities.
- **Modify the random-walk scale prior**: Adjust `sigma ~ normal(0, 1)`
  to allow more or less smoothness in the logit-hazard trajectory.
- **Use a different primary distribution**: Pass any function with a
  registered CDF to `dprimary` to use the analytic convolution.
