---
title: "Fitting distributions using primarycensorseddist and cmdstan"
description: "A guide on how to use primarycensoreddist with Stan for Bayesian inference of epidemiological delay distributions."
author: Sam Abbott
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting distributions using primarycensorseddist and cmdstan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What are we going to do in this Vignette

In this vignette, we'll demonstrate how to use `primarycensoreddist` in conjunction with Stan for Bayesian inference of epidemiological delay distributions. We'll cover the following key points:

1. Simulating censored delay distribution data
2. Fitting a naive model using cmdstan
3. Evaluating the naive model's performance
4. Fitting an improved model using `primarycensoreddist` functionality
5. Comparing the `primarycensoreddist` model's performance to the naive model

## What might I need to know before starting

This vignette builds on the concepts introduced in the [Getting Started with primarycensoreddist](primarycensoreddist.html) vignette and assumes familiarity with using Stan tools as covered in the [How to use primarycensoreddist with Stan](using-stan-tools.html) vignette.

## Packages used in this vignette

Alongside the `primarycensoreddist` package we will use the `cmdstanr` package for interfacing with cmdstan and the `tidybayes` package for posterior analysis. We will also use the `ggplot2` package for plotting.

```{r setup, message = FALSE}
library(primarycensoreddist)
library(cmdstanr)
library(tidybayes)
library(ggplot2)
```

# Simulating censored and truncated delay distribution data

We'll start by simulating some censored and truncated delay distribution data. We'll use the `rprimarycensoreddist` function (actually we will use the `rpcens ` alias for brevity).

```{r sample-lognormal}
set.seed(123) # For reproducibility

# Define the true distribution parameters
n <- 1000
meanlog <- 1.5
sdlog <- 0.75

# Generate varying pwindow, swindow, and obs_time lengths
pwindows <- sample(1:5, n, replace = TRUE)
swindows <- sample(1:5, n, replace = TRUE)
obs_times <- sample(8:12, n, replace = TRUE)

# Function to generate a single sample
generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rlnorm,
    meanlog = meanlog, sdlog = sdlog,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

# Generate samples
samples <- mapply(generate_sample, pwindows, swindows, obs_times)

# Create initial data frame
delay_data <- data.frame(
  pwindow = pwindows,
  swindow = swindows,
  obs_time = obs_times,
  observed_delay = samples
)

head(delay_data)

# Aggregate to unique combinations and count occurrences
# Aggregate to unique combinations and count occurrences
delay_counts <- delay_data |>
  summarise(n = n(), .by = c(pwindow, swindow, obs_time, observed_delay))

head(delay_counts)

# Compare the samples with and without secondary censoring to the true
# distribution
# Calculate empirical CDF
empirical_cdf <- ecdf(samples)

# Create a sequence of x values for the theoretical CDF
x_seq <- seq(min(samples), max(samples), length.out = 100)

# Calculate theoretical CDF
theoretical_cdf <- plnorm(x_seq, meanlog = meanlog, sdlog = sdlog)

# Create a long format data frame for plotting
cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(c("Observed", "Theoretical"), each = length(x_seq))
)

# Plot
ggplot(cdf_data, aes(x = x, y = probability, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c(Observed = "#4292C6", Theoretical = "#252525")
  ) +
  geom_vline(
    aes(xintercept = mean(samples), color = "Observed"),
    linetype = "dashed", linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = exp(meanlog + sdlog^2 / 2), color = "Theoretical"),
    linetype = "dashed", linewidth = 1
  ) +
  labs(
    title = "Comparison of Observed vs Theoretical CDF",
    x = "Delay",
    y = "Cumulative Probability",
    color = "CDF Type"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```
