---
title: "Fitting distributions using primarycensored and fitdistrplus"
description: "A guide on how to fit distributions using primarycensored and fitdistrplus."
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting distributions using primarycensored and fitdistrplus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What you will learn

By the end of this vignette, you will be able to:
- Understand the bias introduced by ignoring primary censoring and truncation when fitting delay distributions
- Use `fitdistdoublecens()` to fit distributions that properly account for primary censoring, secondary censoring, and truncation
- Understand when MLE-based fitting with fitdistrplus is appropriate versus Bayesian approaches with Stan

## How you might adapt this vignette

This vignette uses simulated data, but you can adapt it for your own work:
- **Replace simulation with real data**: Swap out the simulated `delay_data` with your own observations of delays between primary and secondary events
- **Change the distribution**: Replace `"gamma"` with other distributions like `"lnorm"`, `"weibull"`, or `"norm"` depending on what fits your data
- **Vary the censoring windows**: Adjust `pwindow` and `swindow` to match your data's actual primary and secondary censoring intervals
- **Handle different truncation times**: Use varying values in the `D` column if your observations have different maximum observable delays

## What are we going to do in this vignette

In this vignette, we'll demonstrate how to use `primarycensored` in conjunction with `fitdistrplus` for fitting distributions. We'll cover the following key points:

1. Simulating censored delay distribution data
2. Fitting a naive model using `fitdistrplus`
3. Evaluating the naive model's performance
4. Fitting an improved model using `primarycensored` functionality
5. Comparing the `primarycensored` model's performance to the naive model

## What might I need to know before starting

This vignette assumes some familiarity with the `fitdistrplus` package. If you are not familiar with it then you might want to start with the [Introduction to `fitdistrplus`](https://cran.r-project.org/web/packages/fitdistrplus/vignettes/fitdistrplus_vignette.html) vignette.

## How does this vignette differ from fitting distributions with Stan vignette

This vignette is similar to the `vignette("fitting-dists-with-stan")` vignette in that it shows how to fit a distribution using `primarycensored`. However, here we use maximum likelihood estimation (MLE) to fit the distribution, rather than MCMC. In some settings this may result in a faster fit, but in other settings especially when the data is complex, MCMC may be more reliable. The major benefit of the `fitdistrplus` approach is that we don't need to install additional software (Stan) to fit the distribution. **Note that rather than returning credible intervals, the `fitdistrplus` package returns standard errors and confidence intervals.**

## Packages used in this vignette

Alongside the `primarycensored` package we will use the `fitdistrplus` package for fitting distributions. We will also use the `ggplot2` package for plotting and `dplyr` for data manipulation.

```{r setup, message = FALSE}
library(primarycensored)
library(fitdistrplus)
library(ggplot2)
library(dplyr)
```

# Simulating censored and truncated delay distribution data

We'll start by simulating some censored and truncated delay distribution data. We'll use the `rprimarycensored` function (actually we will use the `rpcens ` alias for brevity).

```{r sample-lognormal}
set.seed(123) # For reproducibility

# Define the number of samples to generate
n <- 1000

# Define the true distribution parameters
shape <- 1.77 # This gives a mean of 4 and sd of 3 for a gamma distribution
rate <- 0.44

# Generate fixed pwindow, swindow, and obs_time
pwindows <- rep(1, n)
swindows <- rep(1, n)
obs_times <- sample(8:10, n, replace = TRUE)

# Function to generate a single sample
generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rgamma,
    shape = shape, rate = rate,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

# Generate samples
samples <- mapply(generate_sample, pwindows, swindows, obs_times)

# Create initial data frame
delay_data <- data.frame(
  delay = samples,
  delay_upper = samples + swindows,
  pwindow = pwindows,
  relative_obs_time = obs_times
)

head(delay_data)

# Compare the samples with and without secondary censoring to the true
# distribution
# Calculate empirical CDF
empirical_cdf <- ecdf(samples)

# Create a sequence of x values for the theoretical CDF
x_seq <- seq(0, 10, length.out = 100)

# Calculate theoretical CDF
theoretical_cdf <- pgamma(x_seq, shape = shape, rate = rate)

# Create a long format data frame for plotting
cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(c("Observed", "Theoretical"), each = length(x_seq)),
  stringsAsFactors = FALSE
)

# Plot
ggplot(cdf_data, aes(x = x, y = probability, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c(Observed = "#4292C6", Theoretical = "#252525")
  ) +
  geom_vline(
    aes(xintercept = mean(samples), color = "Observed"),
    linetype = "dashed", linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = shape / rate, color = "Theoretical"),
    linetype = "dashed", linewidth = 1
  ) +
  labs(
    title = "Comparison of Observed vs Theoretical CDF",
    x = "Delay",
    y = "Cumulative Probability",
    color = "CDF Type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  coord_cartesian(xlim = c(0, 10)) # Set x-axis limit to match truncation
```

In this figure you can see the impact of truncation and censoring as the observed distribution has a much lower mean (the vertical dashed blue line) than the true/theoretical distribution (the vertical dashed black line). Our modelling aim is to recover the true parameters of the theoretical distribution from the observed distribution (i.e. recover the black lines from the blue lines).

# Fitting a naive model using `fitdistrplus`

We first fit a naive model using the `fitdistcens()` function. This function is designed to handle secondary censored data but does not handle primary censoring or truncation without extension.

```{r fit-naive-model}
fit <- delay_data |>
  dplyr::select(left = delay, right = delay_upper) |>
  fitdistcens(
    distr = "gamma",
    start = list(shape = 1, rate = 1)
  )

summary(fit)
```

We see that the naive model has fit poorly due to the primary censoring and right truncation in the data.

# Fitting an improved model using `primarycensored` and `fitdistrplus`

We'll now fit an improved model using the `primarycensored` package. There are two approaches:

1. **Recommended**: Use `fitdistdoublecens()` - a convenience wrapper that handles all the complexity for you
2. **Advanced/Educational**: Define custom distribution functions manually (shown first for understanding)

## Understanding the approach (advanced)

This section shows how to manually define custom distribution functions. **You don't need to do this yourself** - `fitdistdoublecens()` handles it automatically. This is included to help you understand what's happening under the hood.

To fit using `fitdistrplus`, we need to define custom distribution functions that account for primary censoring and truncation.
Rather than using `fitdistcens` we use `fitdist` because our functions are handling the censoring themselves.
Note that in this custom implementation for simplicity we are filtering to use only data with the same `obs_time` rather than handling varying observation times.
This means we're using a subset of our simulated data for the estimation.

```{r fit-improved-model}
# Define custom distribution functions using primarycensored
# The try catch is required by fitdistrplus
dpcens_gamma <- function(x, shape, rate) {
  result <- tryCatch(
    {
      dprimarycensored(
        x, pgamma,
        shape = shape, rate = rate,
        pwindow = 1, swindow = 1, D = 8
      )
    },
    error = function(e) {
      rep(NaN, length(x))
    }
  )
  return(result)
}

ppcens_gamma <- function(q, shape, rate) {
  result <- tryCatch(
    {
      pprimarycensored(
        q, pgamma,
        shape = shape, rate = rate,
        dpwindow = 1, D = 8
      )
    },
    error = function(e) {
      rep(NaN, length(q))
    }
  )
  return(result)
}

# Fit the model using fitdistcens with custom gamma distribution
pcens_fit <- delay_data |>
  dplyr::filter(relative_obs_time == 8) |>
  dplyr::pull(delay) |>
  fitdist(
    distr = "pcens_gamma",
    start = list(shape = 1, rate = 1)
  )

summary(pcens_fit)
```

We see good agreement between the true and estimated parameters but with higher standard errors due to using a subset of the data.

## Using `fitdistdoublecens()` (recommended)

Rather than defining custom functions manually, `primarycensored` provides the `fitdistdoublecens()` wrapper function that handles everything automatically. This is the **recommended approach** for most users.

Key advantages of `fitdistdoublecens()`:
- **No custom functions needed**: Just specify the distribution name (e.g., `"gamma"`, `"lnorm"`, `"weibull"`)
- **Handles varying observation windows**: Supports different truncation times (`D`) and primary windows (`pwindow`) across observations
- **Familiar interface**: Uses column names similar to `fitdistcens()` (`left` and `right` for censoring bounds)
- **Supports mixed censoring intervals**: Different secondary censoring windows across observations are handled automatically

```{r primarycensored-fitdistdoublecens}
# Fit using the recommended fitdistdoublecens() wrapper
fitdistdoublecens_fit <- fitdistdoublecens(
  delay_data,
  distr = "gamma",
  start = list(shape = 1, rate = 1),
  left = "delay",
  right = "delay_upper",
  pwindow = "pwindow",
  D = "relative_obs_time"
)

summary(fitdistdoublecens_fit)
```

## Summary

In this vignette we have shown how to fit a distribution using `primarycensored` in conjunction with `fitdistrplus`. The key takeaways are:

- **Use `fitdistdoublecens()`**: This is the recommended approach for most users. It handles all the complexity of primary censoring, secondary censoring, and truncation automatically.
- **Specify distributions by name**: Pass the base distribution name (e.g., `"gamma"`, `"lnorm"`, `"weibull"`) to the `distr` parameter. The function automatically finds the corresponding `d` and `p` functions.
- **Ignoring censoring causes bias**: The naive model that ignores primary censoring and truncation substantially underestimates the true distribution parameters.

For a more robust Bayesian approach to fitting distributions, see the `vignette("fitting-dists-with-stan")` vignette. For more flexible delay distribution fitting, see the [`epidist`](https://epidist.epinowcast.org) package (which uses `primarycensored` under the hood).
