---
title: "Fitting distributions using primarycensorseddist and cmdstan"
description: "A guide on how to use primarycensoreddist with Stan for Bayesian inference of epidemiological delay distributions."
author: Sam Abbott
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting distributions using primarycensorseddist and cmdstan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What are we going to do in this Vignette

In this vignette, we'll demonstrate how to use `primarycensoreddist` in conjunction with Stan for Bayesian inference of epidemiological delay distributions. We'll cover the following key points:

1. Simulating censored delay distribution data
2. Fitting a naive model using cmdstan
3. Evaluating the naive model's performance
4. Fitting an improved model using `primarycensoreddist` functionality
5. Comparing the `primarycensoreddist` model's performance to the naive model

## What might I need to know before starting

This vignette builds on the concepts introduced in the [Getting Started with primarycensoreddist](primarycensoreddist.html) vignette and assumes familiarity with using Stan tools as covered in the [How to use primarycensoreddist with Stan](using-stan-tools.html) vignette.

## Packages used in this vignette

Alongside the `primarycensoreddist` package we will use the `cmdstanr` package for interfacing with cmdstan. We will also use the `ggplot2` package for plotting and `dplyr` for data manipulation.

```{r setup, message = FALSE}
library(primarycensoreddist)
library(cmdstanr)
library(ggplot2)
library(dplyr)
```

# Simulating censored and truncated delay distribution data

We'll start by simulating some censored and truncated delay distribution data. We'll use the `rprimarycensoreddist` function (actually we will use the `rpcens ` alias for brevity).

```{r sample-lognormal}
set.seed(123) # For reproducibility

# Define the true distribution parameters
n <- 1000
meanlog <- 1.5
sdlog <- 0.75

# Generate varying pwindow, swindow, and obs_time lengths
pwindows <- sample(1, n, replace = TRUE)
swindows <- sample(1, n, replace = TRUE)
obs_times <- 10

# Function to generate a single sample
generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rlnorm,
    meanlog = meanlog, sdlog = sdlog,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

# Generate samples
samples <- mapply(generate_sample, pwindows, swindows, obs_times)

# Create initial data frame
delay_data <- data.frame(
  pwindow = pwindows,
  swindow = swindows,
  obs_time = obs_times,
  observed_delay = samples
)

head(delay_data)

# Aggregate to unique combinations and count occurrences
# Aggregate to unique combinations and count occurrences
delay_counts <- delay_data |>
  summarise(n = n(), .by = c(pwindow, swindow, obs_time, observed_delay))

head(delay_counts)

# Compare the samples with and without secondary censoring to the true
# distribution
# Calculate empirical CDF
empirical_cdf <- ecdf(samples)

# Create a sequence of x values for the theoretical CDF
x_seq <- seq(min(samples), max(samples), length.out = 100)

# Calculate theoretical CDF
theoretical_cdf <- plnorm(x_seq, meanlog = meanlog, sdlog = sdlog)

# Create a long format data frame for plotting
cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(c("Observed", "Theoretical"), each = length(x_seq)),
  stringsAsFactors = FALSE
)

# Plot
ggplot(cdf_data, aes(x = x, y = probability, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c(Observed = "#4292C6", Theoretical = "#252525")
  ) +
  geom_vline(
    aes(xintercept = mean(samples), color = "Observed"),
    linetype = "dashed", linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = exp(meanlog + sdlog^2 / 2), color = "Theoretical"),
    linetype = "dashed", linewidth = 1
  ) +
  labs(
    title = "Comparison of Observed vs Theoretical CDF",
    x = "Delay",
    y = "Cumulative Probability",
    color = "CDF Type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

We've aggregated the data to unique combinations of `pwindow`, `swindow`, and `obs_time` and counted the number of occurrences of each `observed_delay` for each combination. This is the data we will use to fit our model.

# Fitting a naive model using cmdstan

We'll start by fitting a naive model using cmdstan. We'll use the `cmdstanr` package to interface with cmdstan. We define the model in a string and then write it to a file as in the [How to use primarycensoreddist with Stan](using-stan-tools.html) vignette.

```{r naive-model}
writeLines(
  "data {
    int<lower=0> N;  // number of observations
    vector[N] y;     // observed delays
    vector[N] n;     // number of occurrences for each delay
  }
  parameters {
    real mu;
    real<lower=0> sigma;
  }
  model {
    // Priors
    mu ~ normal(1, 1);
    sigma ~ normal(0.5, 1);

    // Likelihood
    target += n .* lognormal_lpdf(y | mu, sigma);
  }",
  con = file.path(tempdir(), "naive_lognormal.stan")
)
```

Now let's compile the model

```{r compile-naive-model}
naive_model <- cmdstan_model(file.path(tempdir(), "naive_lognormal.stan"))
```

and now let's fit the compiled model.

```{r fit-naive-model, message = FALSE}
naive_fit <- naive_model$sample(
  data = list(
    # Add a small constant to avoid log(0)
    y = delay_counts$observed_delay + 1e-6,
    n = delay_counts$n,
    N = nrow(delay_counts)
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
naive_fit
```

We see that the model has converged and the diagnostics look good. However, just from the model posterior summary we see that we might not be very happy with the fit. `mu` is smaller than the target `r meanlog` and `sigma` is larger than the target `r sdlog`.

# Fitting an improved model using primarycensoreddist

We'll now fit an improved model using the `primarycensoreddist` package. The main improvement is that we will use the `primary_censored_dist_lpdf` function to fit the model. This is the Stan version of the `pcens()` function and accounts for the primary and secondary censoring windows as well as the truncation. We encode that our primary distribution is a lognormal distribution by passing 1 as the `dist_id` parameter and that our primary event distribution is uniform by passing 1 as the `primary_dist_id` parameter. See the [Stan documentation](https://primarycensoreddist.epinowcast.org/stan/primary__censored__dist_8stan.html#acc97240dee1bc19e4f02013118f3857d) for more details on the `primary_censored_dist_lpdf` function.

```{r pimarycensoreddist-model}
writeLines(
  "
  functions {
    #include primary_censored_dist.stan
    #include expgrowth.stan
  }
  data {
    int<lower=0> N;  // number of observations
    array[N] int<lower=0> y;     // observed delays
    array[N] int<lower=0> n;     // number of occurrences for each delay
    array[N] int<lower=0> pwindow; // primary censoring window
    array[N] int<lower=0> swindow; // secondary censoring window
    array[N] int<lower=0> D; // maximum delay
  }
  transformed data {
    array[0] real primary_params;
    array[N] real y_upper;
    for (i in 1:N) {
      y_upper[i] = y[i] + swindow[i];
    }
  }
  parameters {
    real mu;
    real<lower=0> sigma;
  }
  model {
    // Priors
    mu ~ normal(1, 1);
    sigma ~ normal(0.5, 0.5);

    // Likelihood
    for (i in 1:N) {
      target += n[i] * primary_censored_dist_lpmf(
        y[i] | 1, {mu, sigma},
        pwindow[i], y_upper[i], D[i],
        1, primary_params
      );
    }
  }",
  con = file.path(tempdir(), "primarycensoreddist_lognormal.stan")
)
```

Now let's compile the model

```{r compile-primarycensoreddist-model, message = FALSE}
primarycensoreddist_model <- cmdstan_model(
  file.path(tempdir(), "primarycensoreddist_lognormal.stan"),
  include_paths = pcd_stan_path()
)
```

Now let's fit the compiled model.

```{r fit-primarycensoreddist-model, message = FALSE}
primarycensoreddist_fit <- primarycensoreddist_model$sample(
  data = list(
    y = delay_counts$observed_delay[3],
    n = delay_counts$n[3],
    pwindow = delay_counts$pwindow[3],
    swindow = delay_counts$swindow[3],
    D = delay_counts$obs_time[3],
    N = 1
  ),
  chains = 4,
  parallel_chains = 4,
  init = list( # we use this to resolve initialisation issues
    list(mu = 1.5, sigma = 0.75),
    list(mu = 1.5, sigma = 0.4),
    list(mu = 1.5, sigma = 0.3),
    list(mu = 1.5, sigma = 0.8)
  ),
  refresh = ifelse(interactive(), 100, 0),
  show_messages = interactive()
)
primarycensoreddist_fit
```

We see that the model has converged and the diagnostics look good. We also see that the posterior means are very near the true parameters and the 90% credible intervals include the true parameters.
