---
title: "Fitting delay distributions with negative support"
description: "Estimating a serial-interval-style distribution that can take negative values, using both fitdistrplus and the bundled cmdstan model."
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting delay distributions with negative support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Some delays of epidemiological interest can be negative.
The serial interval, the time between symptom onset in a primary case and symptom onset in a secondary case, is a common example: pre-symptomatic transmission means the secondary case can develop symptoms before the primary case, giving a negative observed serial interval.
Distributions on the non-negative reals (lognormal, gamma, Weibull) cannot represent this and the package's analytical solutions all assume non-negative support.

In this vignette we estimate the parameters of a logistic-distributed serial interval from doubly-censored, right-truncated samples that include negative observations.
We use [`fitdistdoublecens()`] for a quick MLE fit and [`pcd_cmdstan_model()`] for the Bayesian fit.
The logistic is symmetric with slightly heavier tails than the normal, has support on the whole real line, and ships in base R as `dlogis`/`plogis`/`qlogis`/`rlogis`.
On the Stan side it dispatches to `dist_id = 17` and is integrated via the ODE path (no analytical solution is registered for it).

## Packages used in this vignette

```{r setup, message = FALSE}
library(primarycensored)
library(ggplot2)
library(dplyr)
```

# Simulating censored serial interval data

We pick a logistic with location `2` and scale `1` and impose a finite lower truncation `L = -3`: in our worked example we observed serial intervals only from pairs whose secondary case had symptom onset no more than three days before the primary.
A finite `L < 0` is the case the Stan side could not previously handle and is the headline addition here.
Setting `L = -Inf` (the package default) would also work and would correspond to "no lower truncation".

```{r simulate}
set.seed(1)
n <- 2000
location_true <- 2
scale_true <- 1
lower_truncation <- -3

pwindows <- sample.int(2, n, replace = TRUE)
swindows <- sample.int(2, n, replace = TRUE)
obs_times <- sample(10:14, n, replace = TRUE)

generate_sample <- function(pwindow, swindow, obs_time) {
  rprimarycensored(
    1, rlogis,
    location = location_true, scale = scale_true,
    pwindow = pwindow, swindow = swindow, D = obs_time, L = lower_truncation
  )
}
samples <- mapply(generate_sample, pwindows, swindows, obs_times)

delay_data <- data.frame(
  pwindow = pwindows,
  swindow = swindows,
  obs_time = obs_times,
  start_relative_obs_time = lower_truncation,
  observed_delay = samples,
  observed_delay_upper = pmin(obs_times, samples + swindows)
)

cat(sprintf(
  paste0(
    "Observed serial intervals span [%s, %s] with %d negative values ",
    "out of %d.\n"
  ),
  min(samples), max(samples), sum(samples < 0), n
))
```

The simulated data contains a meaningful number of negative observed delays.
A non-negative delay distribution would have no way to assign probability to those rows.

We aggregate the data on its unique combinations of integer-valued covariates so the likelihood is computed once per unique observation rather than once per sample.

```{r aggregate}
delay_counts <- delay_data |>
  summarise(
    n = n(),
    .by = c(
      pwindow, obs_time, start_relative_obs_time,
      observed_delay, observed_delay_upper
    )
  )

head(delay_counts)
```

# MLE fit with `fitdistdoublecens()`

`fitdistdoublecens()` wraps `fitdistrplus::fitdistcens()` so we can pass any distribution name for which `dlogis`/`plogis`/`qlogis` are available.
We pass our left-truncation point `L = -3` via the `L` column and the upper interval via the `right` column.

```{r fitdistdoublecens-fit}
fitdist_data <- mutate(
  delay_data,
  left = observed_delay,
  right = observed_delay_upper,
  L = start_relative_obs_time,
  D = obs_time
)

mle_fit <- fitdistdoublecens(
  fitdist_data,
  distr = "logis",
  start = list(location = 1, scale = 1),
  pwindow = "pwindow",
  L = "L",
  D = "D"
)

summary(mle_fit)
```

The MLE point estimates land close to the true `location = 2`, `scale = 1`.

# Bayesian fit with `pcd_cmdstan_model()`

Because the logistic has support on the real line, observed delays in the data may be negative; the data block now allows this and a `transformed data` check rejects rows where `d[i] < L[i]` or `d_upper[i] > D[i]`.
We pass our left-truncation point through the `start_relative_obs_time` column.

```{r pcd-cmdstan-fit, message = FALSE}
pcd_data <- pcd_as_stan_data(
  delay_counts,
  delay = "observed_delay",
  delay_upper = "observed_delay_upper",
  relative_obs_time = "obs_time",
  start_relative_obs_time = "start_relative_obs_time",
  dist_id = pcd_stan_dist_id("logistic", "delay"),
  primary_id = pcd_stan_dist_id("uniform", "primary"),
  # Lower bound for `scale` is set slightly above zero rather than at zero so
  # the leapfrog integrator cannot underflow `scale = exp(unconstrained)` to
  # exactly 0, which would trip `logistic_lcdf`'s positivity check during
  # warmup.
  param_bounds = list(lower = c(-Inf, 0.01), upper = c(Inf, Inf)),
  primary_param_bounds = list(lower = numeric(0), upper = numeric(0)),
  priors = list(location = c(2, 1), scale = c(1, 0.5)),
  primary_priors = list(location = numeric(0), scale = numeric(0))
)

pcd_model <- pcd_cmdstan_model()
pcd_fit <- pcd_model$sample(
  data = pcd_data,
  chains = 2,
  parallel_chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)

pcd_fit$summary("params")
```

The posterior 90% credible intervals contain the true parameters and the chains mix well (`rhat` close to 1).

# Comparing estimates

```{r compare-plot, fig.cap = "Posterior density (cmdstan) and MLE point estimate against the simulation truth.", fig.width = 7, fig.height = 3}
draws <- pcd_fit$draws("params", format = "data.frame")
mle_pt <- coef(mle_fit)

plot_data <- bind_rows(
  data.frame(
    parameter = "location",
    value = draws[["params[1]"]],
    stringsAsFactors = FALSE
  ),
  data.frame(
    parameter = "scale",
    value = draws[["params[2]"]],
    stringsAsFactors = FALSE
  )
)

truth <- data.frame(
  parameter = c("location", "scale"),
  truth = c(location_true, scale_true),
  mle = c(mle_pt[["location"]], mle_pt[["scale"]]),
  stringsAsFactors = FALSE
)

ggplot(plot_data, aes(x = value)) +
  geom_density(fill = "#4292C6", alpha = 0.5) +
  geom_vline(data = truth, aes(xintercept = truth), linetype = "dashed") +
  geom_vline(data = truth, aes(xintercept = mle), linetype = "dotted") +
  facet_wrap(~parameter, scales = "free") +
  labs(
    x = "Parameter value",
    y = "Posterior density",
    caption = "Dashed = simulation truth, dotted = fitdistdoublecens MLE."
  ) +
  theme_minimal()
```

Both methods recover the simulation truth.
The MLE returns a point estimate and asymptotic standard errors; the Bayesian fit returns the full posterior at the cost of longer runtime.

# Other distributions with support on the reals

The same workflow extends to any signed-support delay that has a `d`/`p`/`q` triple in R and a Stan dispatch.
`pcd_distributions` and `pcd_stan_dist_id()` document the registered options; the negative-support choices currently available on the Stan side are:

- **Logistic** (`dist_id = 17`, used here): symmetric, slightly heavier tails than the normal, in base R.
- **Cauchy** (`dist_id = 12`): symmetric, very heavy tails. Useful for serial intervals where extreme negative values cannot be ruled out. In base R as `dcauchy`/`pcauchy`/`qcauchy`/`rcauchy`.
- **Gumbel** (`dist_id = 15`): right-skewed, supports the reals. A natural choice when the right tail is markedly heavier than the left, but not in base R; you would either define `dgumbel`/`pgumbel`/`qgumbel`/`rgumbel` in your script (the standard formulae are short) or attach a package such as `extraDistr`.

The `dist_id` upper bound in `pcens_model.stan` is currently capped at `17`, so distributions with higher IDs (e.g. normal at `18`) are not yet available through `pcd_cmdstan_model()`.
You can still use them for the R-side `fitdistdoublecens()` fit.

# Adapting this vignette

- **Real serial-interval data**: replace the simulated `delay_data` with an empirical table of `(observed_delay, observed_delay_upper, pwindow, obs_time)` rows.
- **Add a finite `L`**: pass a negative or positive value via the `start_relative_obs_time` column to truncate the distribution from below at `L`.
- **Different distribution**: swap `rlogis`/`"logis"`/`"logistic"` for any of the alternatives listed above (or any base-R / package-supplied distribution with a matching `d`/`p`/`q`/`r` set on the R side).
