---
title: "Fitting distributions using primarycensorseddist and fitdistrplus"
description: "A guide on how to fit distributions using primarycensored and fitdistrplus."
author: Sam Abbott
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting distributions using primarycensored and fitdistrplus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## What are we going to do in this Vignette

In this vignette, we'll demonstrate how to use `primarycensored` in conjunction with `fitdistrplus` for fitting distributions. We'll cover the following key points:

1. Simulating censored delay distribution data
2. Fitting a naive model using `fitdistrplus`
3. Evaluating the naive model's performance
4. Fitting an improved model using `primarycensored` functionality
5. Comparing the `primarycensored` model's performance to the naive model

## What might I need to know before starting

This vignette assumes some familiarity with the `fitdistrplus` package. If you are not familiar with it then you might want to start with the [Introduction to `fitdistrplus`](https://cran.r-project.org/web/packages/fitdistrplus/vignettes/fitdistrplus_vignette.html) vignette.

## Packages used in this vignette

Alongside the `primarycensored` package we will use the `fitdistrplus` package for fitting distributions. We will also use the `ggplot2` package for plotting and `dplyr` for data manipulation.

```{r setup, message = FALSE}
library(primarycensored)
library(fitdistrplus)
library(ggplot2)
library(dplyr)
```

# Simulating censored and truncated delay distribution data

We'll start by simulating some censored and truncated delay distribution data. We'll use the `rprimarycensored` function (actually we will use the `rpcens ` alias for brevity).

```{r sample-lognormal}
set.seed(123) # For reproducibility

# Define the true distribution parameters
n <- 1000
shape <- 1.77 # This gives a mean of 4 and sd of 3 for a gamma distribution
rate <- 0.44

# Generate fixed pwindow, swindow, and obs_time
pwindows <- rep(1, n)
swindows <- rep(1, n)
obs_times <- rep(8, n) # Truncation at 8

# Function to generate a single sample
generate_sample <- function(pwindow, swindow, obs_time) {
  rpcens(
    1, rgamma,
    shape = shape, rate = rate,
    pwindow = pwindow, swindow = swindow, D = obs_time
  )
}

# Generate samples
samples <- mapply(generate_sample, pwindows, swindows, obs_times)

# Create initial data frame
delay_data <- data.frame(
  pwindow = pwindows,
  swindow = swindows,
  obs_time = obs_times,
  observed_delay = samples,
  observed_delay_upper = samples + swindows
)

head(delay_data)

# Compare the samples with and without secondary censoring to the true
# distribution
# Calculate empirical CDF
empirical_cdf <- ecdf(samples)

# Create a sequence of x values for the theoretical CDF
x_seq <- seq(0, 8, length.out = 100)

# Calculate theoretical CDF
theoretical_cdf <- pgamma(x_seq, shape = shape, rate = rate)

# Create a long format data frame for plotting
cdf_data <- data.frame(
  x = rep(x_seq, 2),
  probability = c(empirical_cdf(x_seq), theoretical_cdf),
  type = rep(c("Observed", "Theoretical"), each = length(x_seq)),
  stringsAsFactors = FALSE
)

# Plot
ggplot(cdf_data, aes(x = x, y = probability, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c(Observed = "#4292C6", Theoretical = "#252525")
  ) +
  geom_vline(
    aes(xintercept = mean(samples), color = "Observed"),
    linetype = "dashed", linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = shape / rate, color = "Theoretical"),
    linetype = "dashed", linewidth = 1
  ) +
  labs(
    title = "Comparison of Observed vs Theoretical CDF",
    x = "Delay",
    y = "Cumulative Probability",
    color = "CDF Type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  coord_cartesian(xlim = c(0, 8)) # Set x-axis limit to match truncation
```

# Fitting a naive model using `fitdistrplus`

We first fit a naive model using the `fitdistcens()` function. This function is designed to handle secondary censored data but does not handle primary censoring or truncation without extension.

```{r fit-naive-model}
fit <- delay_data |>
  dplyr::select(left = observed_delay, right = observed_delay_upper) |>
  fitdistcens(
    distr = "gamma",
    start = list(shape = 1, rate = 1)
  )

summary(fit)
```

We see that the naive model has fit poorly due to the primary censoring and right truncation in the data.

# Fitting an improved model using `primarycensored` and `fitdistrplus`

We'll now fit an improved model using the `primarycensored` package. To do this we need to define the custom distribution functions using the `primarycensored` package that are required by `fitdistrplus`. Rather than using `fitdistcens` we use `fitdist` because our functions are handling the censoring themselves.

```{r fit-improved-model}
# Define custom distribution functions using primarycensored
# The try catch is required by fitdistrplus
dpcens_gamma <- function(x, shape, rate) {
  result <- tryCatch(
    {
      dprimarycensored(
        x, pgamma,
        shape = shape, rate = rate,
        pwindow = 1, swindow = 1, D = 8
      )
    },
    error = function(e) {
      rep(NaN, length(x))
    }
  )
  return(result)
}

ppcens_gamma <- function(q, shape, rate) {
  result <- tryCatch(
    {
      pprimarycensored(
        q, pgamma,
        shape = shape, rate = rate,
        pwindow = 1, D = 8
      )
    },
    error = function(e) {
      rep(NaN, length(q))
    }
  )
  return(result)
}

# Fit the model using fitdistcens with custom gamma distribution
pcens_fit <- samples |>
  fitdist(
    distr = "pcens_gamma",
    start = list(shape = 1, rate = 1)
  )

summary(pcens_fit)
```

We see very good agreement between the true and estimated parameters.

Rather than using `fitdist()` directly `primarycensored` provides a wrapper function `fitdistdoublecens()` that can be used to estimate double censored and truncated data. A bonus of this approach is we can specify our data using the `fitdistcens` `left` and `right` formulation and support mixed secondary censoring intervals.

```{r primarycensored-fitdistdoublecens}
fitdistdoublecens_fit <- delay_data |>
  dplyr::select(left = observed_delay, right = observed_delay_upper) |>
  fitdistdoublecens(
    distr = "gamma",
    start = list(shape = 1, rate = 1),
    D = 8, pwindow = 1
  )

summary(fitdistdoublecens_fit)
```


**Note** Currently this functionality is limited to a single pwindow, and observation time for all data. If this functionality is of interest then please open an issue on the `primarycensored` GitHub page.
