---
title: "How to use primarycensored with Stan"
description: "A guide on how to use primarycensored with Stan."
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{How to use primarycensored with Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction

## What are we going to do in this vignette

In this vignette, we'll explore how to use the `primarycensored` package in your Stan modelling workflow. We'll cover the following key points:

1. Introduction to Stan and its relevance for our analysis
2. Overview of the packages we'll be using
3. How to access and use Stan functions provided by `primarycensored`
4. Methods for integrating these Stan functions into your workflow

If you are instead interested in fitting a delay distribution using `primarycensored` in R see the `vignette("fitting-dists-with-stan")` vignette or [`epidist`](https://epidist.epinowcast.org) package (which uses `primarycensored` under the hood).

## What is Stan and why are we using it

Stan is a probabilistic programming language for statistical inference. It provides a flexible and efficient platform for Bayesian modeling and is widely used in various fields of data science and statistics. In this vignette, we'll use Stan in conjunction with `primarycensored` to perform Bayesian inference on censored data.

For more information on Stan:

- [Stan's official website](https://mc-stan.org/)
- [Stan documentation](https://mc-stan.org/users/documentation/)
- [Stan forums](https://discourse.mc-stan.org/) for community support and discussions

## Packages used in this vignette

Alongside the `primarycensored` package we will use the `cmdstanr` package for interfacing with cmdstan.

```{r setup, message = FALSE}
library(primarycensored)
library(cmdstanr)
```

# Using Stan code in primarycensored

`primarycensored` includes a set of Stan functions that mirror the R functions in `primarycensored`. Documentation for these functions can be found [here](https://primarycensored.epinowcast.org/stan/). We support a range of approaches to integrate this Stan code into your workflow.

## Checking available Stan functions using `pcd_stan_functions()`

Aside from reading the documentation it is also possible to list the available Stan functions using a helper function directly in R.

```{r}
pcd_stan_functions()
```

## Accessing Stan functions

Stan functions are accessed using the `pcd_load_stan_functions()` function.
This function takes the name of the function as an argument and returns the function as a string.
It can additionally write the functions to a file and wrap them in a `functions{}` block.

```{r}
pcd_load_stan_functions("primarycensored_lpmf")
```

### Including dependencies automatically

Many Stan functions in `primarycensored` depend on other functions.
For example, `primarycensored_lpmf` calls `primarycensored_lcdf`, which in turn may call analytical or ODE-based implementations.
When using these functions in your own Stan models, you need all the dependencies to be available.

The `dependencies` argument automatically resolves and includes all required functions in the correct order (dependencies before the functions that use them):
```{r}
pcd_load_stan_functions("primarycensored_lpmf", dependencies = TRUE)
```

This is useful when you want to extract a self-contained set of Stan functions for use in another project.

### Exploring function dependencies

To understand which functions a particular Stan function depends on, use `pcd_stan_function_deps()`:

```{r}
pcd_stan_function_deps("primarycensored_lpmf")
```

The result is ordered so that dependencies come before the functions that use them, with the requested function last.
This can help you understand the structure of the package's Stan code or decide which functions you need to include.

## Linking the Stan functions to your workflow

### Writing functions to a file

One option for using Stan functions is to write them to a file and then compile them using `cmdstanr`. This is a good approach as it means that once the functions are written they can be used in the same way as any other stan functions you might use. The downside is that it may mean more work keeping up to date with changes to the functions. We can do this using the `pcd_load_stan_functions()` function.

```{r}
expgrowth_rng_file <- file.path(tempdir(), "expgrowth_rng.stan")
exp_model <- pcd_load_stan_functions(
  "expgrowth_rng",
  write_to_file = TRUE,
  output_file = expgrowth_rng_file,
  wrap_in_block = TRUE
)
```

This can now be compiled and used in the same way as any other `cmdstanr` model.

```{r}
model <- cmdstan_model(expgrowth_rng_file)
model
```

Alternatively, you could use `#include expgrowth_rng.stan` in a stan file functions block to include the function along with the path to that file as with any other stan file (see [here](https://mc-stan.org/cmdstanr/reference/model-method-compile.html)).

### Including the functions directly via `include_paths`

Rather than writing the functions to a file it is also possible to include them directly in the stan file using the `include_paths` argument to `cmdstan_model()`. This is useful if you don't to clutter your model with the stan code from `primarycensored` and want automatic updating of the functions. To demonstrate we will first write a small model has has `expgrowth.stan` in its include paths (rather than writing it to a file and then including it). The first step is find the file and path for the `expgrowth_rng` function.

```{r}
pcd_stan_files("expgrowth_rng")
```

With that done we now write stan wrapper model.

```{r}
expgrowth_stan_file <- file.path(tempdir(), "expgrowth.stan")
writeLines(
  text = c(
    "functions {",
    "#include expgrowth.stan",
    "}",
    "generated quantities {",
    "  real y = expgrowth_rng(0, 1, 0.4);",
    "}"
  ),
  con = expgrowth_stan_file
)
```

We can now use this file to compile a model. **Note** that we need to include the path to the `primarycensored` Stan functions using the `include_paths` argument to `cmdstan_model()`.

```{r}
model <- cmdstan_model(expgrowth_stan_file, include_paths = pcd_stan_path())
model
```

We can then sample from the model (we set `fixed_param = TRUE` here as our toy example doesn't require MCMC sampling).

```{r}
samples <- model$sample(chains = 1, fixed_param = TRUE)
samples
```

## Using Stan functions directly in R

Whilst it is possible to use Stan functions directly in R it is not recommended for most use cases (use the R functions in `primarycensored` instead). However, it can be useful to understand what is going on under the hood or for exploration (indeed we use this internally in `primarycensored` to check our functions against the R implementations). To do this we use the `expose_functions()` method on our already compiled model. **This can take some time (~30 seconds) to compile all of the functions.**

```{r, message = FALSE}
model$expose_functions(global = TRUE)
```

We can now use the function in R. Note that this may get slightly more complicated if our stan function depends on other stan functions (i.e. you need to have those included in your compiled model as well).

```{r}
expgrowth_rng(0, 1, 0.4)
```

## Summary

In this vignette we have shown approaches for using the Stan functions provided by `primarycensored` in your Stan modelling workflow. We have also shown how to use the `expose_functions()` method to access the Stan functions directly in R for exploration and testing.
